### 2021-05-25 concurrency/docs/theory/theory.tex
finish the proof sketch of QueryIter

@@

\newcommand{\Inv}{Inv}
\newcommand{\inv}{inv}
\newcommand{\inlinehoare}[1]{{\color{blue}\left\{#1\right\}}}

Let's use a struct $\rwlock := \{\exc: \Pointer, \rc: \Pointer\}$. Our implementation will be straightforward:

\begin{align*}
  \AcquireExc(\rwlock) ~\eqdef~
      & \langloop~\until~\CAS(\rwlock.\exc,0,1);\\
      & \langloop~\until~(!\rwlock.\rc) = 0\\
      \\
  \AcquireShared(\rwlock) ~\eqdef~
      & \langloop~\until~(\\
        & ~~~~ (\langloop~\until~!\rwlock.\exc);\\
        & ~~~~ \AtomicAdd(\rwlock.\rc, 1);\\
        & ~~~~ \langif~ (!\rwlock.\exc) = 0 ~\langthen\\
        & ~~~~~~~~ \true\\
        & ~~~~ \langelse\\
        & ~~~~~~~~ \AtomicAdd(\rwlock.\rc, -1); \false\\
      )\\
      \\
  %\AcquireExcSetBit(\rwlock) ~\eqdef~
  %    & \CAS(\rwlock.\exc,\false,\true)\\
  %    \\
  %\AcquireExcCheckRc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\rc) = 0\\
  %    \\
  %\AcquireSharedIncRc(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, 1)\\
  %    \\
  %\AcquireSharedAbort(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, -1)\\
  %    \\
  %\AcquireSharedCheckExc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\exc) = \false\\
  %    \\
  \ReleaseExc(\rwlock) ~\eqdef~
      & \rwlock.\exc := 0\\
      \\
  \ReleaseShared(\rwlock) ~\eqdef~
      & \AtomicAdd(\rwlock.\rc, -1)
\end{align*}
We would like to prove the following specifications, for some predicate $\Inv(\rwlock,\alpha,\gamma,\inv)$. (Here, $\inv$ is the user-defined predicate on the ghost state stored by the $\rwlock$.)

@@

\end{document}

>>>>

\newcommand{\BigAsterisk}{\mathop{\scalebox{3}{\raisebox{-0.2ex}{$\ast$}}}}%

\DeclareMathOperator{\RangeFilledWithOtherKeys}{RangeFilledWithOtherKeys}
\DeclareMathOperator{\SlotEmpty}{SlotEmpty}
\DeclareMathOperator{\SlotHasKey}{SlotHasKey}

@@

\newcommand{\Inv}{Inv}
\newcommand{\inv}{inv}
\newcommand{\HTInv}{HTInv}
\newcommand{\Heap}{Heap}
\newcommand{\htinv}{htinv}
\newcommand{\inlinehoare}[1]{{\color{blue}\left\{#1\right\}}}
\newcommand{\inlinehoarel}[1]{{\color{blue}\left\{#1\right.}}
\newcommand{\inlinehoarer}[1]{{\color{blue}\left.#1\right\}}}

Let's use a struct $\rwlock : \{\exc: \Pointer, \rc: \Pointer\}$. Our implementation will be straightforward:

\begin{align*}
  \AcquireExc(\rwlock) ~\eqdef~
      & \langloop~\until~\CAS(\rwlock.\exc,0,1);\\
      & \langloop~\until~(!\rwlock.\rc) = 0\\
      \\
  \AcquireShared(\rwlock) ~\eqdef~
      & \langloop~\until~(\\
        & ~~~~ (\langloop~\until~!\rwlock.\exc);\\
        & ~~~~ \AtomicAdd(\rwlock.\rc, 1);\\
        & ~~~~ \langif~ (!\rwlock.\exc) = 0 ~\langthen\\
        & ~~~~~~~~ \true\\
        & ~~~~ \langelse\\
        & ~~~~~~~~ \AtomicAdd(\rwlock.\rc, -1); \false\\
      )\\
      \\
  %\AcquireExcSetBit(\rwlock) ~\eqdef~
  %    & \CAS(\rwlock.\exc,\false,\true)\\
  %    \\
  %\AcquireExcCheckRc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\rc) = 0\\
  %    \\
  %\AcquireSharedIncRc(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, 1)\\
  %    \\
  %\AcquireSharedAbort(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, -1)\\
  %    \\
  %\AcquireSharedCheckExc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\exc) = \false\\
  %    \\
  \ReleaseExc(\rwlock) ~\eqdef~
      & \rwlock.\exc \langassign 0\\
      \\
  \ReleaseShared(\rwlock) ~\eqdef~
      & \AtomicAdd(\rwlock.\rc, -1)
\end{align*}
We would like to prove the following specifications, for some predicate $\Inv(\rwlock,\alpha,\gamma,\inv)$. (Here, $\inv$ is the user-defined predicate on the ghost state stored by the $\rwlock$.)

@@

\newpage

\textbf{A hash-table logic}

\newcommand{\hashto}{\stackrel{\text{map}}{\hookrightarrow}}
\newcommand{\entryto}{\stackrel{\text{linear}}{\hookrightarrow}}
\newcommand{\Some}{\text{Some}}
\newcommand{\None}{\text{None}}

\newcommand{\Query}{\text{Query}}
\newcommand{\QueryIter}{\text{QueryIter}}
\newcommand{\Update}{\text{Update}}
\newcommand{\UpdateIter}{\text{UpdateIter}}
\newcommand{\Remove}{\text{Remove}}
\newcommand{\RemoveIter}{\text{RemoveIter}}
\newcommand{\NotInRange}{\text{NotInRange}}

\newcommand{\hash}{\ensuremath{hash}}
\newcommand{\rec}{\textbf{rec}}

%We will derive a TPCM (a ``normal'' one, not a refinement) to represent a hash table.
We'll want to support the following specs for a linear-probing hash table.

\begin{align*}
\inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma} &~ \Query(ht,k) ~ \inlinehoare{v.~ \HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma \land v = v_0}\\
\inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma} &~ \Update(ht,k, v) ~ \inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto \Some~v}^\gamma}\\
\inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma} &~ \Remove(ht,k) ~ \inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto \None}^\gamma}
\end{align*}

We assume a function $\hash : Keys \to \mathbb{N}$. We'll let $ht : \{rwlocks: rwlock~list, table: ((Key,Value)~option)~list\}$.

\begin{align*}
  \Query(ht,k) ~\eqdef~ & \QueryIter(ht,k,\hash(k))\\
  \QueryIter(ht,k,i) ~\eqdef~ & 
      \AcquireShared(ht.rwlocks[i]);\\
      &\langlet r = (\langmatch~ !ht.table[i] ~\langwith\\
      &~~~~ |~ \None \Rightarrow \None\\
      &~~~~ |~ \Some~(k_1,v_1) \Rightarrow
          \langif~ k = k_1 ~\langthen~ Some(v_1) ~\langelse~ \QueryIter(ht,k,i+1)\\
      &\langend) ~\langin\\
      &\ReleaseShared(ht.rwlocks[i]);\\
      &r\\
      \\
  \Update(ht,k,v) ~\eqdef~ & \UpdateIter(ht,k,v,\hash(k))\\
  \UpdateIter(ht,k,v,i) ~\eqdef~ & 
      \AcquireExc(ht.rwlocks[i]);\\
      &\langlet r = (\langmatch~ !ht.table[i] ~\langwith\\
      &~~~~ |~ \None \Rightarrow ht.table[i] \langassign (k,v)\\
      &~~~~ |~ \Some~(k_1,v_1) \Rightarrow
          \langif~ k = k_1 ~\langthen~ ht.table[i] \langassign (k,v) ~\langelse~ \UpdateIter(ht,k,v,i+1)\\
      &\langend) ~\langin\\
      &\ReleaseExc(ht.rwlocks[i]);\\
      &r
\end{align*}

\newpage

We'll define a TPCM resource for our hash table (that is, a ``normal'' TPCM, not a refinement). It will have two kinds of elements, $k \hashto v$ (where $v : Value~option$)
and $i \entryto c$, where $c : (Key,Value)~option$.

More precisely, we'll let an element of $HT$ be a pair of maps $(map, linear)$ where
\begin{align*}
k \hashto v ~\eqdef~ ([k := v], [])\\
i \entryto c ~\eqdef~ ([], [i := c])
\end{align*}
$a \cdot b$ is given by pairwise map union, and is undefined $(\bot)$ should either map intersect. For example, we always have $(k \hashto v_1) \cdot (k \hashto v_2) = \bot$
and $(i \entryto c_1) \cdot (i \entryto c_2) = \bot$. (Thus these are very similar to the usual points-to assertions $\pointsto$.

Allowable operations will be encoded into this logic. For example, we will have all the following.
\begin{align*}
  \RangeFilledWithOtherKeys(e, k, a, b) &~\eqdef~ \forall i .~ a \le i < b \implies i \in e.linear \land \exists k_1,v_1 .~ entries[i] = \Some (k_1,v_1) \land k_1 \ne k\\
  %\SlotEmpty(e, a) &~\eqdef~ e.entries[a] = \None\\
  %\SlotHasKey(e, a, k) &~\eqdef~ \exists v.~ e.entries[a] = \Some(k,v)\\
\end{align*}
\emph{Update}.
\begin{align*}
  %&\hash(k) \le u ~\land~ \left(\forall i.~ \hash(k) \le i < u \implies i \in entries \land \exists k_1,v_1 .~ entries[i] = \Some (k_1,v_1) \land k_1 \ne k\right)\\
  %&~~~~\land~ \left(entries[u] = \None ~\lor~ \exists v_1 .~ entries[u] = \Some (k, v) \right)\\
  %&~~~~\implies entries \cdot (k \hashto v_0) ~\mapsto~ entries[u \langassign \Some~(k,v)]  \cdot (k \hashto v)
  &\hash(k) \le u ~\land~ \RangeFilledWithOtherKeys(e, k, \hash(k), u)\\
      %~\land~ \left(\SlotEmpty(e, u) \lor \SlotHasKey(e, a, k)\right)\\
      &~~~~\implies e \cdot (u \entryto \None) \cdot (k \hashto v_0) ~\mapsto~
                    e \cdot (u \entryto \Some~(k, v)) \cdot (k \hashto v)\\
  \\
  &\hash(k) \le u ~\land~ \RangeFilledWithOtherKeys(e, k, \hash(k), u)\\
      %~\land~ \left(\SlotEmpty(e, u) \lor \SlotHasKey(e, a, k)\right)\\
      &~~~~\implies e \cdot (u \entryto \Some~(k, v_1)) \cdot (k \hashto v_0) ~\mapsto~
                    e \cdot (u \entryto \Some~(k, v)) \cdot (k \hashto v)
\end{align*}

\emph{Query}.

The found case is easy: a single entry $(k,v)$ proves that the hash table maps $k$ to $v$.
\[
  \left((i \entryto (k,v_0)) \cdot (k \hashto v)\right) \ne \bot \implies v = \Some(v_0)
\]
The not-found case requires, like the Update case, a check that the key isn't in any of the cells:
\begin{align*}
%&\left(\forall i.~ \hash(k) \le i < u \implies i \in entries \land \exists k_1,v_1 .~ entries[i] = \Some (k_1,v_1) \land k_1 \ne k\right)\\
%  &~~~~\land~ \left(entries[u] = \None\right)\\
%  &~~~~\implies entries \cdot (k \hashto v) \ne \bot \implies v = \None
&\hash(k) \le u \land \RangeFilledWithOtherKeys(e, k, \hash(k), u)
  \implies e \cdot (u \entryto \None) \cdot (k \hashto v) \ne \bot \implies v = \None
\end{align*}

We'll define $\HTInv$ as follows:
\begin{align*}
\HTInv(ht, \gamma) ~\eqdef~& \exists\alpha.~ \BigAsterisk_i \Inv\left(ht.rwlocks[i], \alpha, \gamma\times\Heap, \htinv(i, ht.table[i])\right)\\
\htinv(i, \ell) ~\eqdef~& \lambda x .~ \exists c .~ x = (i \entryto c, \ell \pointsto c)
\end{align*}
In other words, the RwLocks will be arranged so that we can use the lock at index $i$ to obtain, possibly in read-only ``shared'' mode,
some ghost state $(i \entryto c, ht.table[i] \pointsto c)$.

\newpage

Now, we prove $\QueryIter$. (To avoid clutter, we silently skolemize existential
variables as they are introduced.)
The program begins with an established borrow at lifetime $\kappa$.
We then acquire a shared lock, obtaining the $\SharedGuard$.
We then borrow that, and from that, obtain borrowed state for the points-to assertion to read a cell from memory,
along with the corresponding $\entryto$ state.
The new borrow has lifetime $\kappa_0$. We intersect them to get $\kappa_1 = \kappa \ltIntersect \kappa_0$. We downgrade
all the borrows to $\kappa_1$, which we will use for the recursive call. The sketch below shows step-by-step the process
of introducing the new borrow, expiring it, and returning the lock (we handle the $\langmatch$ statement on the next page).
We leave $\HTInv(ht,\gamma)$ implicit.
\begin{align*}
  &\inlinehoare{\active(\kappa) * \borrow{\kappa}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i)}\\
  &\AcquireShared(ht.rwlocks[i]);\\
  &\inlinehoare{\active(\kappa) * \borrow{\kappa}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i)
      * \SharedGuard\left( (i \entryto c, ht.table[i] \pointsto c) \right) }\\
  &\inlinehoarel{\active(\kappa) * \borrow{\kappa}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i)}\\
  &\inlinehoarer{~~~~
      * \active(\kappa_0) * \borrowed\left(\kappa_0, \SharedGuard\left( (i \entryto c, ht.table[i] \pointsto c) \right) \right)
      * \borrow{\kappa_0}{\SharedGuard\left( (i \entryto c, ht.table[i] \pointsto c) \right) }}\\
  &\inlinehoarel{\active(\kappa) * \borrow{\kappa}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i) * \borrow{\kappa_0}{\dboxed{i\entryto c}^\gamma} * \borrow{\kappa_0}{ht.table[i] \pointsto c}}\\
  &\inlinehoarer{~~~~
      * \active(\kappa_0) * \borrowed\left(\kappa_0, \SharedGuard\left( (i \entryto c, ht.table[i] \pointsto c) \right) \right) }\\
  &\inlinehoarel{\active(\kappa_1) * \borrow{\kappa_1}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i) * \borrow{\kappa_1}{\dboxed{i\entryto c}^\gamma} * \borrow{\kappa_1}{ht.table[i] \pointsto c}}\\
  &\inlinehoarer{~~~~
      * \borrowed\left(\kappa_0, \SharedGuard\left( (i \entryto c, ht.table[i] \pointsto c) \right) \right) }\\
  &\langlet r = (\langmatch~ !ht.table[i] ~\langwith\\
  &~~~~ |~ \None \Rightarrow \None\\
  &~~~~ |~ \Some~(k_1,v_1) \Rightarrow
      \langif~ k = k_1 ~\langthen~ \Some(v_1) ~\langelse~ \QueryIter(ht,k,i+1)\\
  &\langend) ~\langin\\
  &\inlinehoare{\active(\kappa_1)
      * \dboxed{k \hashto v_0}^\gamma \land r = v_0
      * \borrowed\left(\kappa_0, \SharedGuard\left( (i \entryto c, ht.table[i] \pointsto c) \right) \right) }\\
  &\inlinehoare{\active(\kappa)
      * \dboxed{k \hashto v_0}^\gamma \land r = v_0
      * \SharedGuard\left( (i \entryto c, ht.table[i] \pointsto c) \right) }\\
  &\ReleaseShared(ht.rwlocks[i]);\\
  &\inlinehoare{\active(\kappa) * \dboxed{k \hashto v_0}^\gamma \land r = v_0}\\
  &r\\
  &\inlinehoare{v.~ \active(\kappa) * \dboxed{k \hashto v_0}^\gamma \land v = v_0}
\end{align*}

\newpage

The ``interesting'' logic is all in the match statement. Depending on the value of $c$, we either use our borrowed $\dboxed{i \entryto c}^\gamma$ to justify the result,
or we combine it with $e$ for the recursive call (extending the range in the $\RangeFilledWithOtherKeys$ predicate to $i+1$).
We let $e' = e \cdot (i \entryto \Some~(k_1,v_1))$.

\begin{align*}
  &\inlinehoare{\active(\kappa_1) * \borrow{\kappa_1}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i) * \borrow{\kappa_1}{\dboxed{i\entryto c}^\gamma} * \borrow{\kappa_1}{ht.table[i] \pointsto c}}\\
  &\langlet r = (\langmatch~ !ht.table[i] ~\langwith\\
  &~~~~ |~ \None \Rightarrow \\
  &~~~~~~~~\inlinehoare{\active(\kappa_1) * \borrow{\kappa_1}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i) * \borrow{\kappa_1}{\dboxed{i\entryto \None}^\gamma}}\\
  &~~~~~~~~\inlinehoare{\active(\kappa_1)
      * \dboxed{k \hashto v_0}^\gamma \land v_0 = \None}\\
  &~~~~~~~~\None\\
  &~~~~ |~ \Some~(k_1,v_1) \Rightarrow \\
  &~~~~~~~~\langif~ k = k_1 ~\langthen~ \\
  &~~~~~~~~~~~~\inlinehoare{\active(\kappa_1) * \borrow{\kappa_1}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i) * \borrow{\kappa_1}{\dboxed{i\entryto \Some~(k,v_1)}^\gamma}}\\
  &~~~~~~~~~~~~\inlinehoare{\active(\kappa_1)
      * \dboxed{k \hashto v_0}^\gamma \land v_0 = \Some(v_1)}\\
  &~~~~~~~~~~~~\Some(v_1)\\
  &~~~~~~~~~\langelse~\\
  &~~~~~~~~~~~~\inlinehoare{\active(\kappa_1) * \borrow{\kappa_1}{\dboxed{e}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e, k, \hash(k), i) * \borrow{\kappa_1}{\dboxed{i\entryto \Some~(k_1,v_1)}^\gamma} \land k\ne k_1}\\
  &~~~~~~~~~~~~\inlinehoare{\active(\kappa_1) * \borrow{\kappa_1}{\dboxed{e'}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land \RangeFilledWithOtherKeys(e', k, \hash(k), i+1)}\\
  &~~~~~~~~~~~~~~~~~~~~~~~~~~~~\text{\color{blue} (where $e' = e \cdot (i \entryto \Some~(k_1,v_1))$)}\\
  &~~~~~~~~~~~~\QueryIter(ht,k,i+1)\\
  &~~~~~~~~~~~~\inlinehoare{v.~ \active(\kappa_1) * \dboxed{k \hashto v_0}^\gamma \land v = v_0}\\
  &\langend) ~\langin\\
  &\inlinehoare{\active(\kappa_1)
      * \dboxed{k \hashto v_0}^\gamma \land r = v_0}
\end{align*}
\end{document}
